---
title: "final project scRNA"
output: html_document
author: "Emily Johnson" 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  

## Libraries
``` {r eval = TRUE}
library(dplyr)
library(Seurat)
library(edgeR)
library(limma)
library(Glimma)
library(gplots)
library(org.Mm.eg.db)
library(RColorBrewer)
```

## Reading in data 
Reading in count data (not raw). There are 5 types of count data: iPSCs, ESCs, embryos, mouse adult dopaminergic cells, and mouse embryos. 
```{r eval = TRUE} 
ips <- read.delim("GSE76381_iPSMoleculeCounts.cef.txt", stringsAsFactors = FALSE)
embryo <- read.delim("GSE76381_EmbryoMoleculeCounts.cef.txt", stringsAsFactors = FALSE)
es<- read.delim("GSE76381_ESMoleculeCounts.cef.txt", stringsAsFactors = FALSE)
mouse.adult.da <- read.delim("GSE76381_MouseAdultDAMoleculeCounts.cef.txt", stringsAsFactors = FALSE)
mouse.embryo <- read.delim("GSE76381_MouseEmbryoMoleculeCounts.cef.txt", stringsAsFactors = FALSE)

ips.seurat <- CreateSeuratObject(counts = ips, project = "BMIN", min.cells = 0, min.features = 0)
embryo.seurat <- CreateSeuratObject(counts = embryo, project = "BMIN", min.cells = 0, min.features = 0)
es.seurat <- CreateSeuratObject(counts = es, project = "BMIN", min.cells = 0, min.features = 0)
mouse.adult.da.seurat <- CreateSeuratObject(counts = mouse.adult.da, project = "BMIN", min.cells = 0, min.features = 0)
mouse.embryo.seurat <- CreateSeuratObject(counts = mouse.embryo, project = "BMIN", min.cells = 0, min.features = 0)

```

## Formatting Data
renaming each column to cell IDs
``` {r eval = TRUE}
names(ips) <- ips[1, ]
ips <- ips[-1,]
ips.counts <- ips[-(1:3),-2]

names(embryo) <- embryo[1, ]
embryo <- embryo[-1,]

names(ips) <- ips[1, ]
ips <- ips[-1,]

names(es) <- es[1, ]
es <- es[-1,]

names(mouse.adult.da) <- mouse.adult.da[1, ]
mouse.adult.da <- mouse.adult.da[-1,]

names(mouse.embryo) <- mouse.embryo[1, ]
mouse.embryo <- mouse.embryo[-1,]

```

## Cleaning Data 
Finding expression levels 
``` {r eval = TRUE}

counts_per_cell <- Matrix::colSums(ips)
counts_per_gene <- Matrix::rowSums(ips)
genes_per_cell <- Matrix::colSums(ips>0)

```

## Filtering Low Expression 
Group each matrix by cell type, and look for expression for cell types 


## Testing Practicum 17 read in 
``` {r eval = TRUE}
library(GEOquery)
library(oligo)
library(limma)
library(viridis)
library(hgu133plus2.db)
library(ggplot2)
library(gplots)

getGEOSuppFiles("GSE76381")
untar("./GSE76381/GSE76381_RAW.tar", exdir = "./GSE76381/data")

celFiles <- list.celfiles("./GSE76381/data", full.names = TRUE, listGzipped = TRUE)
raw.data <- read.celfiles(celFiles)

pData(raw.data)


gse <- getGEO("GSE76381", GSEMatrix = TRUE)
show(gse)

```


